Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-11-18 text/ms.tex
Incorporate some of Jon's comments and add TODOs

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, heavy-tailed distributions have been discovered in
time-series of marine microbe abundance \citep{segura2013} and time to
extinction for experimental waterflea populations \citep{drake2014}. Despite
these anecotes, as far as we can tell there have been two systematic surveys
for ecological black swans: one with North American breeding birds and the
other with the Global Population Dynamics Database, which uncovered little
clear evidence for them \citep{keitt1998,allen2001,halley2002}. Nevertheless
there are methodological challenges to their detection
\citep{allen2001,ward2007}.

@@

Here we develop a framework for identifying ecological black swans in the
process noise of population dynamics, i.e.\ the stochastic jumps in abundance
from one time step to the next. Our framework quantifies both the magnitude and
probability of heavy-tailed dynamics, allows for a range of population dynamic
models, can incorporate observation uncertainty, and can be easily applied to
abundance time series. We apply our model to \NPops\ populations and estimate
the frequency and strength of black-swan events in population dynamics across
taxonomy; verify some causes of black-swan events; and attempt to identify
characteristics of time series or intrinsic life-history characteristics that
are associated with heavy-tailed dynamics.

@@

We found strong, but rare, evidence for black-swan population dynamics. By
defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
10$, our main Gompertz model found evidence for heavy tails most frequently
for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
orders. Accounting for time series length and partially pooling inference
across taxonomic class and order with a multilevel model, there was stronger
evidence for black swans in insect populations than is visually apparent in
Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median
probability of heavy tails were insect orders---however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).

@@

We examined all time series with published explanations of why the black-swan
events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
majority (\percBSDown \%) of documented events were downward black swans and
involved a series of unfortunate events. For example, a shortage of nest sites
for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the
United Kingdom reduced population productivity and made the population more
vulnerable when a red tide hit in 1968 \citep{potts1980}. The population
experienced a black-swan downswing, but the new availability of nest sites
resulted in a rapid population upswing \citep{potts1980}. As another example,
an interaction between environmental- and predation-mediated population cycles
are thought to have caused a downward black-swan event for a water vole
(\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans
were the result of a sequence of extreme climate events on their own. For
instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated
with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in
the United Kingdom \citep{stafford1971}. Our analysis finds that the last event
was a combination of two black-swan events in a row and it took the population
three times longer to recover than predicted \citep{stafford1971}.
\textit{(TODO: substitute a parasite example?)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\%
of ecological time series. Black swans were usually downward events and were
detected more frequently in longer time series and in populations with a
smaller magnitude of process noise. Black swans were not associated with
density dependence, population growth rate, or lifespan. In verified cases,
black-swan events were often a result of the interaction between elements of
extreme climate, predators and parasite cycles, and strong changes in human
pressures. Our empirical results, sensitivity analyses, and simulation tests
suggest that estimating the tail shape of process noise is a viable method of
detecting black-swan population dynamics and if anything will underestimate the
probability of black-swan events. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics---particularly as the
frequency and magnitude of climatic extremes increase over the next century
\citep{easterling2000,ipcc2012}.

>>>>

Black swans are statistically improbable events that nonetheless occur---often
with profound consequences. While extremes in the physical environment, such
as monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations
buffer or suffer from such extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in
ecological process noise. We apply our model to \NPops~time series from around
the world across \NOrders~taxonomic orders and seven classes. We find strong
evidence of black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\% of populations: most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), and insects
(\InsectaRangePerc\%). Black swans were predominantly (\percBSDown \%)
downward events and were not explained by any life-history covariates, but
tended to be driven by external perturbations such as climate, severe winters,
predator or parasite cycles, and the combined effects of multiple factors.
Extreme events were more frequently detected for populations with longer time
series and lower levels of process noise; for shorter and noisier time series,
our simulations suggested black-swan dynamics are often not identifiable as
such. The presence of black swans in population dynamics highlights the
importance of developing robust conservation and management
strategies---particularly as the frequency and magnitude of climate extremes
increase over the next century.

@@

There is compelling anecdotal evidence for ecological black swans, but
systematic evidence across taxa has been elusive. A survey of ecologists
indicated that surprising outcomes of field experiments are far more common
than we assume \citep{doak2008}, and events such as the global invasion of
Argentine ants and the mutation of viruses to infect new hosts could be
considered black swans \citep{nunez2012}. In fact, anecdotes of population
catastrophes are numerous and catastrophes may be the most important element
affecting population persistence \citep{mangel1994}. For marine mammal
populations, we have compelling evidence of catastrophes \citep{gerber2001,
  ward2007}, and recently, time-series of marine microbe abundance
\citep{segura2013} and time to extinction for experimental waterflea
populations \citep{drake2014} have been found to have heavier tailed
distributions than the normal. Despite these anecdotes, as far as we can tell
there have been two systematic surveys for ecological black swans: one with
North American breeding birds and the other with the Global Population
Dynamics Database, which uncovered little clear evidence for them
\citep{keitt1998,allen2001,halley2002}. Nevertheless there are methodological
challenges to their detection \citep{allen2001,ward2007}.

@@

%TODO: this paragraph is written as if geared towards a methods journal. I suggest starting with a statement of general aim rather than method.
Here we develop a framework for identifying ecological black swans in the
process noise of population dynamics, i.e.\ the stochastic jumps in abundance
from one time step to the next. Our framework quantifies both the magnitude and
probability of heavy-tailed dynamics, allows for a range of population dynamic
models, can incorporate observation uncertainty, and can be easily applied to
abundance time series. We apply our model to \NPops\ populations and estimate
the frequency and strength of black-swan events in population dynamics across
taxonomy; verify some causes of black-swan events; and attempt to identify
characteristics of time series or intrinsic life-history characteristics that
are associated with heavy-tailed dynamics.

@@

%TODO: where did this threshold come from?  (nu < 10)
We found strong, but rare, evidence for black-swan population dynamics. By
defining black-swan dynamics as a greater than $0.5$ probability that $\nu <
10$, our main Gompertz model found evidence for heavy tails most frequently
for birds (\birdPH\%) followed by mammals (\mammalsPH\%), and insects
(\insectsPH\%) (Fig.~\ref{fig:nu-coefs}, Table~\ref{tab:causes-supp}). Black
swans were taxonomically widespread, occurring in \POrdersHeavy\% of taxonomic
orders. Accounting for time series length and partially pooling inference
across taxonomic class and order with a multilevel model, there was stronger
evidence for black swans in insect populations than is visually apparent in
Fig.~\ref{fig:nu-coefs}---four of 10 orders with the highest median
probability of heavy tails were insect orders---however, there was
considerable uncertainty in these estimates (Fig.~\ref{fig:posteriors}a).
%TODO: these multilevel modelling results (and fig 3, 4) are out of order

@@

We examined all time series with published explanations of why the black-swan
events occurred (Tables~\ref{tab:sparks} and \ref{tab:causes-supp}). The
majority (\percBSDown \%) of documented events were downward black swans and
%TODO: what else would cause a black swan event? How defined "unfortunate"?
involved a series of unfortunate events. For example, a shortage of nest sites
%TODO how about just calling it a piscivorous waterfowl or something like that
for the shag long-necked bird (\textit{Phalacrocorax aristotelis}) in the
%TODO why did this make the population more vulnerable? Seems like it wouldn't make it more vulnerable I guess I am trying to figure out if these are additive or synergistic. I am guessing you want to avoid those terms, but I might just be more straight with language and avoid the "more vulnerable" phrase.
United Kingdom reduced population productivity and made the population more
vulnerable when a red tide hit in 1968 \citep{potts1980}. The population
experienced a black-swan downswing, but the new availability of nest sites
resulted in a rapid population upswing \citep{potts1980}. As another example,
% TODO true interaction term or just synchronous events?
an interaction between environmental- and predation-mediated population cycles
are thought to have caused a downward black-swan event for a water vole
(\textit{Arvicola terrestris}) population \citep{saucy1994}. Other black swans
were the result of a sequence of extreme climate events on their own. For
instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated
with black-swan downswings in grey heron (\textit{Ardea cinerea}) abundance in
the United Kingdom \citep{stafford1971}. Our analysis finds that the last event
was a combination of two black-swan events in a row and it took the population
three times longer to recover than predicted \citep{stafford1971}.
\textit{(TODO: substitute a parasite example?)}

@@

We found strong evidence for black swans (heavy-tailed process noise) in 3--5\%
% TODO but the students t is a symmetric distribution is that cool?
of ecological time series. Black swans were usually downward events and were
detected more frequently in longer time series and in populations with a
smaller magnitude of process noise. Black swans were not associated with
density dependence, population growth rate, or lifespan. In verified cases,
% TODO 'interaction' word
black-swan events were often a result of the interaction between elements of
extreme climate, predators and parasite cycles, and strong changes in human
pressures. Our empirical results, sensitivity analyses, and simulation tests
suggest that estimating the tail shape of process noise is a viable method of
detecting black-swan population dynamics and if anything will underestimate the
probability of black-swan events. The presence of black swans highlights the
importance of developing management strategies that detect quickly, respond to,
and are robust to extremes in population dynamics---particularly as the
frequency and magnitude of climatic extremes increase over the next century
\citep{easterling2000,ipcc2012}.

### 2014-11-18 text/sparks.tex
Tweak sparks table caption
Incorporate some of Jon's comments and add TODOs

\caption{Example population dynamic black swans from the Global Population
  Dynamics Database and a description of their causes. Red and blue dots
  indicate downward and upward events that have a $10 \cdot 10^{-4}$ probability or less of
  occurring if the population dynamics were explained by a Gompertz model
  with normally distributed process noise. These populations are a non-random
  sample that we verified in the primary literature.}\\

@@

\includegraphics[width=2cm]{../analysis/sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm
environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

@@

\includegraphics[width=2cm]{../analysis/sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

>>>>

\caption{Example population dynamic black swans from the Global Population
  Dynamics Database and a description of their causes. Red and blue dots
  indicate downward and upward events that have a $10 \cdot 10^{-4}$ probability or less of
  occurring if the population dynamics were explained by a Gompertz model
  with normally distributed process noise. These populations are
  a sample from the heavy-tailed populations we could verify
  (Table~\ref{tab:causes-supp}).}\\

@@

\includegraphics[width=2cm]{../analysis/sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with long-term
environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

@@

\includegraphics[width=2cm]{../analysis/sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was buried under snow caused population crash &
\citep{newton1998} \\
